Learning Auditory Similarities Between Instrumental Playing Techniques

### 2019-12-02 paper/paper.tex
describe hypergraph partitioning

We consider a list of 16 instruments: violin (Vn), viola (Va), cello (Vc), contrabass (Cb), concert harp (Hp), Spanish guitar (Gtr), accordion (Acc), flute (Fl), soprano clarinet (BbCl), alto saxophone (ASax), oboe (Ob), bassoon (Bn), trumpet in C (TpC), French horn (Hn), tenor trombone (TTbn), and bass tuba (BBTb).
Among this list, the former six are strings, the next six are woodwind, and the last four are brass.
Some of these instruments may be temporarily equipped with timbre-altering mutes, such as a rubber sordina on the bridge of a violin or an aluminium ``wah-wah'', also known as \emph{harmon}, inside in the bell of a trumpet.
Once augmented with mutes, the list of 16 instruments grows up to 33.
Furthermore, every instrument, whether equipped with a mute or not, affords a panel of playing techniques ranging in size between 11 (for the accordion) and 41 (for the bass tuba).
In the rest of this paper, we abbreviate instrument--mute--technique by means of the acronym ``IMT''.
One example of IMT is \texttt{TpC+S-ord}, \ie{} trumpet in C with a straight mute played in the ordinary technique.
Another example of IMT is \texttt{Vn-pont}, \ie{} violin without any mute played in the \emph{sul ponticello} (bowing near the bridge) technique.

@@

First, we purposefully disentangle categorical variables (IMTs) from continuous variables (pitch and intensity) in the SOL dataset.
Indeed, under first approximation, the perception of timbre is invariant to pitch and intensity.
Therefore, we select auditory stimuli according to a reference pitch and a reference intensity; in our case, middle C (\ie{} C4) and \emph{mf}. % TODO use pitch spelling package
After this selection, every IMT triplet contains a single acoustic exemplar, regarded as canonical in the following.
The number of canonical stimuli for the entire SOL dataset is equal to 235.
We should note, however, that the proposed pitch--intensity cannot be strictly enforced across all IMTs.
Indeed, as explained above, a fraction of IMTs can only be achieved at restricted values of pitch and intensity parameters, \eg{} pedal tones or key clicks.
Therefore, at a small cost of consistency, we only enforce the pitch--intensity reference (\ie{} C4 and \emph{mf}) when practically feasible, and fall back to other pitches and intensities if necessary.

Secondly, we reduce the number of IMTs in our study by focusing on those which are deemed to be most relevant.
Here, we define the relevance of an IMT as the possibility of imitating it by means of another IMT from a different instrument.
One example of such imitation is the acoustic similarity between slap tonguing in reed instruments and a snap pizzicato in strings instruments.
To collect perceptual ratings of relevance, we recruited two professors in music composition at the Paris Conservatory (CNSMDP), and explicitly gave them our definition of relevance.
Each of them inspected the entire corpus of $235$ IMTs and annotated them in terms of relevance according to a Likert scale with seven ticks.
In this Likert scale, the value 1 (least relevant) denotes that the IMT under consideration has a timbre that is idiosyncratic of the instrument, and that therefore, it is unlikely that humans will pair it with IMTs from other instruments.
Conversely, the value 7 (most relevant) denotes that the IMT under consideration has a strong similarity with IMTs from one or several other instruments.
Once both experts completed their annotation, we retained all IPTs whose average relevance score was judged equal to 3 or higher.
At the end of this stage of pre-screening, the list of IMT reduces from 235 to $N=78$.
Tables 1 and 2, in the Appendix section of this paper, give a full list of these 78 highly relevant IMTs.
It is worth noting that, according to both experts, the timbre of the accordion was judged too idiosyncratic, regardless of playing technique, to be relevant for this experiment.
Indeed, the accordion is the only instrument in the aforementioned list of instrument to have free reeds, keyboard-based actuation, or handheld airflow.

Thirdly, we design a graphical user interface for partitioning a corpus of short audio samples.
The need for such an interface arises from the unscalability of Likert scales in the context of pairwise similarity judgments.
Assuming that similarity is a symmetric quantity, collecting a dense matrix of continuously valued ratings of similarity among a dataset of $N$ items would require $\frac{1}{2}(N^2-N)$ Likert scales.
In the case of $N=78$ IMTs, the task would amount to about 3k horizontal sliders, \ie{} several hours of cumbersome work for the human annotator.
Engaging as many participants as possible in our study called for a more streamlined form of human--computer interaction, even if it sacrificed the availability of continuous ratings.
To this end, we implemented a web application, named Cyberlioz, in which the user can spontaneously listen and arrange sounds into clusters of timbre similarity\footnote{The Web application for efficient audio annotation, as well as the raw anonymized responses of all $31$ participants to our study, is available at: \url{https://soundthings.org/research/cyberlioz/}}.
The name Cyberlioz is a portmanteau between the prefix cyber- and the French composer Hector Berlioz, whose \emph{Treatise on Orchestration} (1844) marked a watershed in the history of European classical music, because of its deliberate focus on the role of timbre as a parameter for musical expression.

@@

In comparison with a traditional Web-based form, Cyberlioz offers a more intuitive and playful user experience, while expediting the obtention of a dense matrix ($N\times N$) of similarity judgments within a moderate duration of $30$ to $60$ minutes for each participant.
In X 2016, we recruited volunteers to use Cyberlioz on their own computers, via a web browser, and equipped with a pair of earphones. % TODO give the exact two months.
We publicized this study on international mailing lists for research in music audio processing, such as X and Y. % TODO give mailing lists
Within two months, $K=31$ subjects accessed Cyberlioz and completed the task. % TODO break down by gender? give age stats? musical experience?
Figure \ref{fig:xp2display}. % TODO

@@

We aggregate the similarity judgments from all $K$ subjects by embedding them into a hypergraph $\mathcal{H}$, that is, a graph whose edges may connect three or more vertices at once.
Specifically, $\mathcal{H}$ contains $N$ vertices, each representing an IMT; and each ``hyperedge'' in $\mathcal{H}$ corresponds to some connected component in one of the graphs $\mathcal{G}_k$'s among all subject $k$.
Then, we convert the hypergraph $\mathcal{H}$ back into a conventional graph $\mathcal{G}_0$ by means of a combinatorial optimization algorithm algorithm known as hypergraph partitioning \cite{kernighan1970efficient,han1997scalable,strehl2002cluster}. % TODO describe the gist of hypergraph partitioning?
While the graphs $\mathcal{G}_1 \ldots \mathcal{G}_K$ encode the subjective similarity judgments of participants $1$ to $K$, the graph $\mathcal{G}_0$ represents a form of consensual judgment that is shared across all participants, while discarding intersubjective variability.
Although the rest of our paper focuses on the consensus $\mathcal{G}_0$, it is worth pointing out that the same technical framework could apply to a single subject $k$, or to a subgroup of the $K=31$ subjects.
This remark emphasizes the potential of our similarity learning method as a customizable tool for visualizing and extrapolating the timbre similarity space of a newcoming subject, after less than one hour of using Cyberlioz.

>>>>

\subsection*{Dataset}
We consider a list of 16 instruments: violin (Vn), viola (Va), cello (Vc), contrabass (Cb), concert harp (Hp), Spanish guitar (Gtr), accordion (Acc), flute (Fl), soprano clarinet (BbCl), alto saxophone (ASax), oboe (Ob), bassoon (Bn), trumpet in C (TpC), French horn (Hn), tenor trombone (TTbn), and bass tuba (BBTb).
Among this list, the former six are strings, the next six are woodwind, and the last four are brass.
Some of these instruments may be temporarily equipped with timbre-altering mutes, such as a rubber sordina on the bridge of a violin or an aluminium ``wah-wah'', also known as \emph{harmon}, inside in the bell of a trumpet.
Once augmented with mutes, the list of 16 instruments grows up to 33.
Furthermore, every instrument, whether equipped with a mute or not, affords a panel of playing techniques ranging in size between 11 (for the accordion) and 41 (for the bass tuba).
In the rest of this paper, we abbreviate instrument--mute--technique by means of the acronym ``IMT''.
One example of IMT is \texttt{TpC+S-ord}, \ie{} trumpet in C with a straight mute played in the ordinary technique.
Another example of IMT is \texttt{Vn-pont}, \ie{} violin without any mute played in the \emph{sul ponticello} (bowing near the bridge) technique.

@@

\subsection*{Disentanglement of factors}
First, we purposefully disentangle categorical variables (IMTs) from continuous variables (pitch and intensity) in the SOL dataset.
Indeed, under first approximation, the perception of timbre is invariant to pitch and intensity.
Therefore, we select auditory stimuli according to a reference pitch and a reference intensity; in our case, middle C (\ie{} C4) and \emph{mf}. % TODO use pitch spelling package
After this selection, every IMT triplet contains a single acoustic exemplar, regarded as canonical in the following.
The number of canonical stimuli for the entire SOL dataset is equal to 235.
We should note, however, that the proposed pitch--intensity cannot be strictly enforced across all IMTs.
Indeed, as explained above, a fraction of IMTs can only be achieved at restricted values of pitch and intensity parameters, \eg{} pedal tones or key clicks.
Therefore, at a small cost of consistency, we only enforce the pitch--intensity reference (\ie{} C4 and \emph{mf}) when practically feasible, and fall back to other pitches and intensities if necessary.

\subsection*{Expert pre-screening}
Secondly, we reduce the number of IMTs in our study by focusing on those which are deemed to be most relevant.
Here, we define the relevance of an IMT as the possibility of imitating it by means of another IMT from a different instrument.
One example of such imitation is the acoustic similarity between slap tonguing in reed instruments and a snap pizzicato in strings instruments.
To collect perceptual ratings of relevance, we recruited two professors in music composition at the Paris Conservatory (CNSMDP), and explicitly gave them our definition of relevance.
Each of them inspected the entire corpus of $235$ IMTs and annotated them in terms of relevance according to a Likert scale with seven ticks.
In this Likert scale, the value 1 (least relevant) denotes that the IMT under consideration has a timbre that is idiosyncratic of the instrument, and that therefore, it is unlikely that humans will pair it with IMTs from other instruments.
Conversely, the value 7 (most relevant) denotes that the IMT under consideration has a strong similarity with IMTs from one or several other instruments.
Once both experts completed their annotation, we retained all IPTs whose average relevance score was judged equal to 3 or higher.
At the end of this stage of pre-screening, the list of IMT reduces from 235 to $N=78$.
Tables 1 and 2, in the Appendix section of this paper, give a full list of these 78 highly relevant IMTs.
It is worth noting that, according to both experts, the timbre of the accordion was judged too idiosyncratic, regardless of playing technique, to be relevant for this experiment.
Indeed, the accordion is the only instrument in the aforementioned list of 
instrument to have free reeds, keyboard-based actuation, or handheld airflow.

\subsection*{Efficient annotation interface}
Thirdly, we design a graphical user interface for partitioning a corpus of short audio samples.
The need for such an interface arises from the unscalability of Likert scales in the context of pairwise similarity judgments.
Assuming that similarity is a symmetric quantity, collecting a dense matrix of continuously valued ratings of similarity among a dataset of $N$ items would require $\frac{1}{2}(N^2-N)$ Likert scales.
In the case of $N=78$ IMTs, the task would amount to about 3k horizontal sliders, \ie{} several hours of cumbersome work for the human annotator.
Engaging as many participants as possible in our study called for a more streamlined form of human--computer interaction, even if it sacrificed the availability of continuous ratings.
To this end, we implemented a web application, named Cyberlioz, in which the user can spontaneously listen and arrange sounds into clusters of timbre similarity\footnote{The Web application for efficient audio annotation, as well as the raw anonymized responses of all $31$ participants to our study, is available at: \url{https://soundthings.org/research/cyberlioz/}}.
The name Cyberlioz is a portmanteau between the prefix cyber- and the French composer Hector Berlioz, whose \emph{Treatise on Orchestration} (1844) marked a watershed in the history of European classical music, because of its deliberate focus on the role of timbre as a parameter for musical expression.

@@

In comparison with a traditional Web-based form, Cyberlioz offers a more intuitive and playful user experience, while expediting the obtention of a dense matrix ($N\times N$) of similarity judgments within a moderate duration of $30$ to $60$ minutes for each participant.
In X 2016, we recruited volunteers to use Cyberlioz on their own computers, via a web browser, and equipped with a pair of earphones. % TODO give the exact two months.
We publicized this study on international mailing lists for research in music audio processing, such as X and Y. % TODO give mailing lists
Within two months, $K=31$ subjects accessed Cyberlioz and completed the task. % TODO break down by gender? give age stats? musical experience?
%Figure \ref{fig:xp2display}. % TODO

@@

We aggregate the similarity judgments from all $K$ subjects by embedding them into a hypergraph $\mathcal{H}$, that is, a graph whose edges may connect three or more vertices at once.
Specifically, $\mathcal{H}$ contains $N$ vertices, each representing an IMT; and each ``hyperedge'' in $\mathcal{H}$ corresponds to some connected component in one of the graphs $\mathcal{G}_k$'s among all subject $k$.
Then, we convert the hypergraph $\mathcal{H}$ back into a conventional graph $\mathcal{G}_0$ by means of a combinatorial optimization algorithm algorithm known as hypergraph partitioning \cite{kernighan1970efficient,han1997scalable,strehl2002cluster}. % TODO describe the gist of hypergraph partitioning?
To construct $G_0$, we select a number of clusters $C=5$ a priori, and run hypergraph partitioning to assign each vertex $i$ to one of these clusters.
Intuitively, hypergraph partitioning optimizes a tradeoff between two objectives: first, balancing the size of all clusters in terms of their respective numbers of vertices; and secondly, keeping most hyperedges enclosed within as few distinct clusters as possible.

While the graphs $\mathcal{G}_1 \ldots \mathcal{G}_K$ encode the subjective similarity judgments of participants $1$ to $K$, the graph $\mathcal{G}_0$ represents a form of consensual judgment that is shared across all participants, while discarding intersubjective variability.
Although the rest of our paper focuses on the consensus $\mathcal{G}_0$, it is worth pointing out that the same technical framework could apply to a single subject $k$, or to a subgroup of the $K=31$ subjects.
This remark emphasizes the potential of our similarity learning method as a customizable tool for visualizing and extrapolating the timbre similarity space of a newcoming subject, after less than one hour of using Cyberlioz.
